# Hotspot maps

Now that we've learned how to work with `ggplot2()`, we can make a hotspot map using many of the same concepts as the graphs we have already done. The concept of determining the geographic locations where crime is most intense - crime hotspots - is very important to criminological theory as well as to the very practical question of where police officers should be deployed.

Here, we will make hotspot maps with our officer_shootings_geocoded crime data. First we need to load the data.

```{r}
load("data/officer_shootings_geocoded.rda")
```


## Creating the maps
We'll be using two packages to make our maps, `ggmap` and `hexbin`, so be sure you have those installed.

```{r, eval=FALSE}
install.packages(c("ggmap", "hexbin"))
```

```{r, cache=FALSE}
library(ggmap)
library(hexbin)
```

Next, we'll need to get a map of officer_shootings_geocoded. Google Maps used to be readily available through `ggmap`, but as of June 2018 the account requirements became more complicated for Google Maps. We will use other map services instead here. Stamen is a design organization that provides maps of several styles. Here we define the bounding box for the city of Philadelphia. This bounding box determines where in the world the map shows, so we want to set it to coorindates that will show the whole of Philadelphia.

Next, we'll need to get a map of officer_shootings_geocoded. Google Maps used to be readily available through `ggmap`, but as of June 2018 the account requirements became more complicated for Google Maps. We will use other map services instead here. Stamen is a design organization that provides maps of several styles. Here we define the bounding box for the City of officer_shootings_geocoded.

```{r, message=FALSE}
philly_map <- ggmap(get_map(c(-75.288486,39.868285,-74.950965,40.138251), source = "stamen"))
philly_map
```

Since we saved the map output into "philly_map" we can reuse this map background for all the maps we're making in this lesson. This saves us time as we don't have to wait to download the map every time. Let's plot the shootings from our data set.

```{r}
philly_map +
  geom_point(aes(x = lon, y = lat),
             data  = officer_shootings_geocoded,
             alpha = 0.5,
             color = "darkred",
             size  = 1)
```

This map is useful because it allows us to easily see where each officer-involved shooting in Philly happened between 2007 and early 2019. There are some limitations though. This shows all shootings in a single map, meaning any time trends are lost - we'll address this at the end of the lesson. While you can see some clusters, it is difficult to see if there are any hot spots indicating areas with especially high or low amounts of shootings. You'll notice that it has a message saying "Removed 2 rows containing missing values (geom_point)." That just means that `ggplot()` automatically deleted 2 rows that didn't contain any lon or lat data. Most data will have a small number of rows missing values, that is nothing to be concerned about unless the number of missing values gets large enough to affect your data.

Let's pause for a moment to think about what a map really is. Below I made a simple scatterplot of our data with one dot per shooting (minus the two without coordinates). Compare this to the map above and you'll see that they are the same except the map has a useful background while the plot has a blank background. That is all static maps are (in a later section we'll learn about interavtive maps), scatterplots of coordinates overlayed on a map background. Basically, they are scatterplots with context. And this context is useful, we can interpret the map to se that there are lots of shootings in West Philly but not around University City, for example, The exact same pattern is present in the scatterlot but without the ability to tell "where" a dot is. 

```{r}
plot(officer_shootings_geocoded$lon, officer_shootings_geocoded$lat, col = "darkred")
```

Let's now try plotting hexagonal bins, which the `hexbin` package let us do. Why hexagons? It turns out that squares tend to have problems in the corners. The precision is low there and our eye tends to get drawn to the parallel lines the square grid makes. Ideally, we want to use a shape that has a small perimeter-to-area ratio. Circles have the smallest perimeter-to-area ratio, but we can't use circles to tile over the map. Hexagons fall in between squares and circles.

Note: when you run the following lines of code, you will get the following message: "Coordinate system already present. Adding new coordinate system, which will replace the existing one." Don't panic, this message is supposed to appear.

```{r}
philly_map +
  coord_cartesian() +
  stat_binhex(aes(x = lon, y = lat),
              bins = 60,
              data = officer_shootings_geocoded)
```

Or try the following. Setting `alpha = 2/4` makes the bins a little more transparent. Try comparing with `alpha = 1/4` or `alpha = 3/4`.

```{r}
philly_map +
  coord_cartesian() +
  stat_binhex(aes(x = lon, y = lat),
              bins  = 60,
              alpha = 0.5,
              data  = officer_shootings_geocoded)
```

The previous map was a default monochromatic color (blue). You can change the color gradient. To see what colors are available, type `colors()`.

```{r}
philly_map +
  coord_cartesian() +
  stat_binhex(aes(x = lon, y = lat),
              bins  = 60,
              alpha = 1,
              data = officer_shootings_geocoded) +
  scale_fill_gradient('Shootings Density',
                      low = "springgreen",
                      high = "darkred")
```

From this map we can see that most areas which has a shooting has only 1-2 shootings in that area (a limitation to this map is it is unclear just how wide these "areas" are). North Philly and North-West Philly has more areas with more shootings, and more shootings overall. But look right in the middle of Center City by City Hall, an area has 8 shootings, the most of any hexagonal. Were there really 8 shootings outside of City Hall? No. Remember those addresses that couldn't be properly located? The coordinates given to those addresses were to City Hall, causing this weird pattern in our map. 

You can also create a map that shows the high crime areas in a way that is akin to elevation on a topographic map. The more concentrated the concentric areas, the higher the crime (or, in the case of a topographic map, the higher the terrain). Note that a plot using `stat_density2d()` requires a lot of memory. That's why it's a good idea to use a sample of cases or a small data set.

```{r}
philly_map +
  stat_density2d(aes(x = lon,y = lat),
                 bins = 5,
                 data = officer_shootings_geocoded,
                 geom = 'density2d',
                 col  = 'white')
```

Let's make a hotspot map with a gradient, red for high crime and green for low crime.

```{r}
philly_map +
  stat_density2d(aes(x = lon, y = lat,
                     fill=..level..,
                     alpha=..level..),
                 data = officer_shootings_geocoded,
                 geom = 'polygon') +
  scale_fill_gradient('Shootings Density',
                      low  = "green",
                      high = "red") +
  scale_alpha(range = c(.4, .75),
              guide = FALSE) +
  guides(fill = guide_colorbar(barwidth = 1.5, barheight = 10))
```

We can use `bins` to control the number of levels of colors plotted.

```{r}
philly_map +
  stat_density2d(aes(x = lon, y = lat,
                     fill = ..level..,
                     alpha = ..level..),
                 bins = 20,
                 data = officer_shootings_geocoded,
                 geom = 'polygon') +
  scale_fill_gradient('Shootings Density',
                      low  = "green",
                      high = "red") +
  scale_alpha(range = c(.4, .75),
              guide = FALSE) +
  guides(fill = guide_colorbar(barwidth = 1.5, barheight = 10)) 
```

Be careful with maps like these. This map is so broad is appears like shootings are ubiquituous across the city. We know from the map showing each shooting as a dot and that there are <500 shootings, that this is  not true. Making maps like this make it easy to mislead the reader, including yourself if you are using maps to better understand your data.

Since our maps show every shooting on the same map, if there are different patterns depending on the year, we can't pick up those patterns. As a check, let's write a for loop that maps each year's data separately. First we need to make a variable with the shooting's year to be able to subset data for just that year. We can use the `year()` function from `lubridate` to do this.

```{r}
library(lubridate)
officer_shootings_geocoded$year <- year(officer_shootings_geocoded$dates)
```

As a basic check, let's just see how many shootings there were each year. 

```{r}
table(officer_shootings_geocoded$year)
```

It looks like the general pattern is that shootings are declining. Now we can map each year. We are looping through each year in the data, and reusing the map code before except instead of the full data, we subset data to be only rows matching the year in that iteration of the loop.  

```{r, results='asis'}
for (year in unique(officer_shootings_geocoded$year)) { 
  print(
    philly_map +
  coord_cartesian() +
  stat_binhex(aes(x = lon, y = lat),
              bins  = 60,
              alpha = 0.75,
              data = officer_shootings_geocoded[officer_shootings_geocoded$year == year, ]) +
  scale_fill_gradient('Shootings Density',
                      low = "springgreen",
                      high = "darkred") +
       ggtitle(paste("Philadelphia Police Officer-Involed Shootings, ", year))
  )
}
```
